`SCPCL001114` - Exploration of gene set based methods for tumor cell assignment

### 2024-07-16 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000752/SCPCL001114_tumor-gene-expression-exploration.Rmd
use two marker genes

# reference files to use
ref_sce_file <- file.path(data_dir, "SCPCP000015", "SCPCS000490", "SCPCL000822_processed.rds")
ref_labels_file <- file.path(module_base, "results", "annotation_tables", "SCPCS000490", "SCPCL000822_tumor-classifications.tsv")
ref_geneset_scores_file <- file.path(module_base, "results", "cnv_annotation", "SCPCS000490", "SCPCL000822_gene-set-scores.tsv")
```

@@

```{r}
# generate classification df to use for plots 
classification_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
  ) |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2, singler_celltype_annotation) |> 
  # join with previous annotations, singler results, and gene set scores 
  dplyr::left_join(auc_results_df, by = "barcodes") |> 
  dplyr::left_join(marker_results_df, by = "barcodes") |> 
  dplyr::left_join(geneset_scores_df, by = "barcodes")

@@

First we will look at a confusion matrix between both methods. 
Here the classification using expression of at least 1 marker gene is the rows and the updated classification based on `SCPCL000822` is the columns.

```{r}
classification_df <- classification_df |> 
  dplyr::mutate(
    marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
  )

# compare using a confusion matrix
caret::confusionMatrix(
    table(
      classification_df$ref_marker_gene_classification,
      classification_df$marker_gene_classification)
  ) 
```

@@

@@

```{r}
method_levels <- c(
  "1 marker gene" = "marker_gene_classification",
  "Ref marker gene" = "ref_marker_gene_classification",
  "AUCell" = "auc_classification", 
  "Consensus" = "consensus_classification"
)

@@

```{r}
# pull out any new classification methods we added 
new_classification_df <- classification_df |> 
  dplyr::select(barcodes, ref_marker_gene_classification, consensus_classification)

@@

annotation <- ComplexHeatmap::columnAnnotation(
  marker_genes = annotation_df$marker_gene_classification,
  ref_marker_genes = annotation_df$ref_marker_gene_classification,
  AUCell = annotation_df$auc_classification, 
  consensus = annotation_df$consensus_classification,
  col = list(marker_genes = plot_colors,
             ref_marker_genes = plot_colors, 
             AUCell = plot_colors,
             consensus = plot_colors))
```

@@

```{r}
# export final TSV with annotations 
classifications_output <- classification_df |> 
  dplyr::select(
    cell_barcode = barcodes,
    marker_gene_classification,
    ref_marker_gene_classification, 
    auc_classification,
    consensus_classification
  )

>>>>

# reference files to use
ref_sce_file <- file.path(data_dir, "SCPCP000015", "SCPCS000490", "SCPCL000822_processed.rds")
ref_labels_file <- file.path(module_base, "results", "annotation_tables", "SCPCS000490", "SCPCL000822_tumor-classifications.tsv.gz")
ref_geneset_scores_file <- file.path(module_base, "results", "cnv_annotation", "SCPCS000490", "SCPCL000822_gene-set-scores.tsv")
```

@@

```{r}
# generate classification df to use for plots 
classification_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
  ) |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2, singler_celltype_annotation) |> 
  # join with previous annotations, singler results, and gene set scores 
  dplyr::left_join(auc_results_df, by = "barcodes") |> 
  dplyr::left_join(marker_results_df, by = "barcodes") |> 
  dplyr::left_join(geneset_scores_df, by = "barcodes") |> 
  # rename since we are going to be looking at multiple marker gene methods 
  dplyr::rename(
    "one_marker_gene_classification" = marker_gene_classification
  )

@@

First we will look at a confusion matrix between both methods. 
Here the updated classification based on `SCPCL000822` is the rows and the classification using expression of at least 1 marker gene is the columns.

```{r}
classification_df <- classification_df |> 
  dplyr::mutate(
    one_marker_gene_classification = forcats::fct_relevel(one_marker_gene_classification, "Tumor")
  )

# compare using a confusion matrix
caret::confusionMatrix(
    table(
      classification_df$ref_marker_gene_classification,
      classification_df$one_marker_gene_classification)
  ) 
```

@@

### Expression of at least two marker genes

Below we will look at classifying cells based on expression of two marker genes rather than one.

```{r}
# get list of tumor cells with > 2 marker genes expressed 
tumor_cells <- plot_markers_df |> 
  dplyr::mutate(gene_detected = as.integer(gene_expression > 0)) |> 
  dplyr::group_by(barcodes) |> 
  dplyr::summarize(total_genes = sum(gene_detected)) |> 
  dplyr::filter(total_genes >= 2) |> 
  dplyr::pull(barcodes)

classification_df <- classification_df |> 
  dplyr::mutate(
    two_marker_gene_classification = dplyr::if_else(
      barcodes %in% tumor_cells,
      "Tumor",
      "Normal"
    ))
```

```{r}
ggplot(classification_df, aes(x = UMAP1, y = UMAP2, color = two_marker_gene_classification)) +
  geom_point(alpha = 0.5, size = 0.5)
```

Now let's compare the annotations obtained using a cutoff determined by `SCPCL000822` to setting the cutoff of 2 marker genes. 
Here the updated classification based on `SCPCL000822` is the rows and the classification using expression of at least 2 marker genes is the columns.

```{r}
classification_df <- classification_df |> 
  dplyr::mutate(
    two_marker_gene_classification = forcats::fct_relevel(two_marker_gene_classification, "Tumor")
  )

# compare using a confusion matrix
caret::confusionMatrix(
    table(
      classification_df$ref_marker_gene_classification,
      classification_df$two_marker_gene_classification)
  ) 
```

It looks like using `SCPCL000822` to define the cutoff results in more cells being called normal than with the strict cutoff of expressing at least 2 marker genes.

@@

```{r}
method_levels <- c(
  "1 marker gene" = "one_marker_gene_classification",
  "2 marker genes" = "two_marker_gene_classification",
  "Ref marker gene" = "ref_marker_gene_classification",
  "AUCell" = "auc_classification", 
  "Consensus" = "consensus_classification"
)

@@

```{r}
# pull out any new classification methods we added 
new_classification_df <- classification_df |> 
  dplyr::select(barcodes, ref_marker_gene_classification, consensus_classification, two_marker_gene_classification)

@@

annotation <- ComplexHeatmap::columnAnnotation(
  one_marker = annotation_df$one_marker_gene_classification,
  two_marker = annotation_df$two_marker_gene_classification,
  ref_marker_genes = annotation_df$ref_marker_gene_classification,
  AUCell = annotation_df$auc_classification, 
  consensus = annotation_df$consensus_classification,
  col = list(one_marker = plot_colors,
             two_marker = plot_colors,
             ref_marker_genes = plot_colors, 
             AUCell = plot_colors,
             consensus = plot_colors))
```

@@

```{r}
# export final TSV with annotations 
classifications_output <- classification_df |> 
  dplyr::select(
    cell_barcode = barcodes,
    one_marker_gene_classification,
    two_marker_gene_classification,
    ref_marker_gene_classification, 
    auc_classification,
    consensus_classification
  )
